Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-01 paper/paper/SPIE/article.tex
WIP on paper

@@

In recent years, RMT analyses have been applied to study brain functioning. The earliest study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) \cite{sebaRandomMatrixAnalysis2003}.
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data \cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

@@

However, these previous studies took largely descriptive / explanatory approaches, noting only
differences in RMT across subgroups and/or conditions, in many cases, with no formal statistical
model or testing of the significance and robustness of these differences. No code was provided to
allow the reproduction of the results, and the extraction of the spectral rigidity and level
variance is non-trivial, and computationally demanding. Computation of these spectral observables
requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}, which is,
unfortunately, a poorly-documented fitting procedure that requires making a number of
difficult-to-justify decisions, and these decisions can dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012,moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to empirical data, there
is also the flexibility introduced by the choices in the complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.

>>>>

With the increase and exapansion of digitial and electronic
Extracting useful features from complex data is a perennial challenge.

@@

Recently, RMT has been applied to study brain functioning. The earliest study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) \cite{sebaRandomMatrixAnalysis2003}.
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data \cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

@@

However, these previous studies took largely descriptive / explanatory approaches, noting only
differences in RMT across subgroups and/or conditions, in many cases, with no formal statistical
model nor formal testing of the significance and robustness of these differences. No code was provided to
allow the reproduction of the results, and the extraction of the spectral rigidity and level
variance is non-trivial, and computationally demanding. Computation of these spectral observables
requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}, which is,
unfortunately, a poorly-documented fitting procedure that requires making a number of
difficult-to-justify decisions, and these decisions can dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012,moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to empirical data, there
is also the flexibility introduced by the choices in the complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.
